Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-26 paper/sections/proof-reconstruction.tex
[ paper ] resolve, nnf, cnf, clausify theorems added.

(φ₁ ⇒ (φ₂ ∧ φ₃))
\begin{equation*}
\scalebox{0.95}{
($\mathcal{D}_1$)
\begin{bprooftree}
\AxiomC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n) ∧ \unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n-1)$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$,}
\end{bprooftree}
}
\end{equation*}

@@

(φ ≡ φ₁ ∧ φ₂). We get a proof for each conjunct and using the introduction rule for conjunction, the result follows.

@@

(φ ≡ φ₁ ⇒ φ₂).
\begin{equation*}
\begin{bprooftree}
\AxiomC{}
\RightLabel{assume φ₁}
\UnaryInfC{$Γ , φ₁ ⊢ φ₁$}
\AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
\RightLabel{weaken φ₁}
\UnaryInfC{$Γ , φ₁ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
\RightLabel{⇒-elim}
\BinaryInfC{$Γ , φ₁ ⊢ \mysplit(φ₂,n-1)$}
\RightLabel{\thmsplit}
\UnaryInfC{$Γ , φ₁ ⊢ φ₂$}
\RightLabel{⇒-intro}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$.}
\end{bprooftree}
\end{equation*}
\vskip 3mm

@@

\begin{theorem}[thm-conjunct]  $Γ ⊢ φ$, and $ψ \in \Prop$, then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}

@@

\begin{lemma}[lem-build-or]
\label{lem:lem-build-or}
$Γ ⊢ φ$, and $ψ\in \Prop$ then $Γ ⊢ build_{∨}(φ)$.
\end{lemma}

@@

Now, with a right-associative proposition,
$strongBuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions.

\begin{align}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &strongBuild_{∨}(φ, ψ) =\\
    &\hspace{3mm}
    \begin{cases}
      factor(build_{∨}(φ₁, ψ) ∨ build_{∨}(φ₂,ψ))  & φ≡ φ₁ ∨ φ₂,\\
      build_{∨}(φ, ψ)  & \text{otherwise.}
    \end{cases}
  \end{split}
\end{align}

\begin{lemma}[lem-sbuild-or]
\label{lem:lem-sbuild-or}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ strongBuild_{∨}(φ, ψ)$.
\end{lemma}

Let see the following example of the idea behind $strongBuild_{∨}$
function; from the disjunction $(r ∨ (q ∨ p))$, we build the disjunction
$((p ∨ q) ∨ r)$.

@@

\begin{theorem}[thm-reorder$_{∨}$] $Γ ⊢ φ$, and $ψ\in\Prop$ then $Γ ⊢ reorder_{∨}(φ, ψ)$, where,
\begin{equation*}
  \label{eq:reorder-or}
  reorder_{∨} (φ, ψ) = strongBuild_{∨}(rdisj(φ), ψ).
\end{equation*}
\end{theorem}
\begin{proof} Use Lemma~\ref{lem:lem-rdisj} and Lemma~\ref{lem:lem-sbuild-or}.
\end{proof}

@@

\begin{lemma}[lem-reorder$_{∧}$]
\label{lem:lem-reorder-and}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder(φ, ψ)$, where,
\begin{align*}
    \begin{split}
      reorder_{∧}(φ, ψ) =
      \begin{cases}
        φ &\text{if }φ ≡ ψ, \\
        ψ₁ ∧ ψ₂ &\text{if } ψ ≡ ψ₁ ∧ ψ₂\text{, }reorder_{∧}(φ, ψ₁) ≡ ψ₁,\\
                &\text{and }reorder_{∧}(φ, ψ₂)≡ ψ₂,\\
        φ       &\text{if } ψ ≡ ψ₁ ∧ ψ₂,\\
        conjunct(φ, ψ) &\text{otherwise.}
      \end{cases}
    \end{split}
\end{align*}

\end{lemma}

\begin{lemma}[lem-conjunct$_{∨}$]
\label{lem:lem-conjunct-or}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,

\begin{align*}
\begin{split}
  \label{eq:conjunct-or-definition}
  conjunct_{∨}(φ, ψ)=
  \begin{cases}
    ψ &\text{if }reorder_{∨}(φ, ψ) ≡ ψ,\\
    ψ &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) = ψ₁,\\
      &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂,\\
    ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ,\\
    ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ,\\
    φ &\text{otherwise.}
  \end{cases}
\end{split}
\end{align*}

\end{lemma}

\begin{theorem}[thm-reorder$_{∧∨}$]
\label{thm:thm-reorder-or-and}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder_{∧∨}(φ, ψ)$, where,

\begin{align}
    \begin{split}
    reorder_{∧∨}(φ, ψ)=
      \begin{cases}
        ψ &\text{if } φ≡ψ,\\
        ψ &\text{if } ψ ≡ ψ₁ ∧ ψ₂, reorder_{∧∨}(φ,ψ₁) ≡ ψ₁,\\
          &\text{and }reorder_{∧∨}(φ, ψ₂) ≡ ψ₂,\\
        φ &\text{if }ψ ≡ ψ₁ ∧ ψ₂,\\
        conjunct_{∨}(φ, ψ) &\text{otherwise.}
      \end{cases}
    \end{split}
\end{align}

\end{theorem}

>>>>

($φ₁ ⇒ (φ₂ ∧ φ₃)$)
\begin{equation*}
\scalebox{0.95}{
($\mathcal{D}_1$)
\begin{bprooftree}
\AxiomC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n) ∧ \unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n-1)$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$,}
\end{bprooftree}
}
\end{equation*}

@@

($φ ≡ φ₁ ∧ φ₂$). We get a proof for each conjunct and using the introduction rule for conjunction, the result follows.

@@

($φ ≡ φ₁ ⇒ φ₂$).
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁$}
  \AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{weaken φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ , φ₁ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₂$}
  \RightLabel{⇒-intro}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$.}
  \end{bprooftree}
\end{equation*}
\vskip 3mm

@@

\begin{theorem}[thm-conjunct]
  \label{thm:thm-conjunct}
  $Γ ⊢ φ$, and $ψ \in \Prop$ , then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}

@@

\begin{lemma}[lem-build$_{∨}$]
\label{lem:lem-build-or}
$Γ ⊢ φ$, and $ψ\in \Prop$ then $Γ ⊢ build_{∨}(φ)$.
\end{lemma}

@@

Now, with a right-associative proposition,
$sbuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions.

\begin{align}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &sbuild_{∨}(φ, ψ) =\\
    &\hspace{3mm}
    \begin{cases}
      factor(build_{∨}(φ₁, ψ) ∨ build_{∨}(φ₂,ψ))  & φ≡ φ₁ ∨ φ₂,\\
      build_{∨}(φ, ψ)  & \text{otherwise.}
    \end{cases}
  \end{split}
\end{align}

\begin{lemma}[lem-sbuild$_{∨}$]
\label{lem:lem-sbuild-or}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ sbuild_{∨}(φ, ψ)$.
\end{lemma}

Let see the following example of the idea behind $sbuild_{∨}$
function; from the disjunction $(r ∨ (q ∨ p))$, we build the disjunction
$((p ∨ q) ∨ r)$.

@@

\begin{theorem}[thm-reorder$_{∨}$]
  \label{thm-reorder-or}
  $Γ ⊢ φ$, and $ψ\in\Prop$ then $Γ ⊢ reorder_{∨}(φ, ψ)$, where,
  \begin{equation*}
    \label{eq:reorder-or}
    reorder_{∨} (φ, ψ) = sbuild_{∨}(rdisj(φ), ψ).
  \end{equation*}
\end{theorem}

\begin{proof} Use Lemma~\ref{lem:lem-rdisj} and Lemma~\ref{lem:lem-sbuild-or}.
\end{proof}

@@

\begin{lemma}[lem-reorder$_{∧}$]
  \label{lem:lem-reorder-and}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder(φ, ψ)$, where,
  \begin{align*}
      \begin{split}
        reorder_{∧}(φ, ψ) =
        \begin{cases}
          φ &\text{if }φ ≡ ψ, \\
          ψ₁ ∧ ψ₂ &\text{if } ψ ≡ ψ₁ ∧ ψ₂\text{, }reorder_{∧}(φ, ψ₁) ≡ ψ₁,\\
                  &\text{and }reorder_{∧}(φ, ψ₂)≡ ψ₂,\\
          φ       &\text{if } ψ ≡ ψ₁ ∧ ψ₂,\\
          conjunct(φ, ψ) &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-conjunct$_{∨}$]
  \label{lem:lem-conjunct-or}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,
  \begin{align*}
  \begin{split}
    \label{eq:conjunct-or-definition}
    conjunct_{∨}(φ, ψ) =
    \begin{cases}
      ψ &\text{if }reorder_{∨}(φ, ψ) ≡ ψ,\\
      ψ &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) = ψ₁,\\
        &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂,\\
      ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ,\\
      ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ,\\
      φ &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

\begin{theorem}[thm-reorder$_{∧∨}$]
  \label{thm:thm-reorder-and-or}

  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder_{∧∨}(φ, ψ)$, where,

  \begin{align}
      \begin{split}
      reorder_{∧∨}(φ, ψ)=
        \begin{cases}
          ψ &\text{if } φ≡ψ,\\
          ψ &\text{if } ψ ≡ ψ₁ ∧ ψ₂, reorder_{∧∨}(φ,ψ₁) ≡ ψ₁,\\
            &\text{and }reorder_{∧∨}(φ, ψ₂) ≡ ψ₂,\\
          φ &\text{if }ψ ≡ ψ₁ ∧ ψ₂,\\
          conjunct_{∨}(φ, ψ) &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{theorem}

% --------------------------------------------------------------------

\subsubsection{Resolve} % (fold)
\label{ssub:resolve}

The \resolve rule is a generalization version for resolution in propositional logic since the formulas are not shaped always as the theorem expect (see Fig.~\ref{fig:metis-inferences}).

Since our approach is mere syntactic, in order to apply such a
theorem, we perform rearrangements inside the formulas to match with
the pattern expected by the resolution theorem.

\begin{lemma}[lem-rsol]
  \label{lem:lem-rsol}
  $Γ ⊢ φ$ then $Γ ⊢ rsol(φ)$, where,
  \begin{align*}
      \begin{split}
        rsol(φ) =
        \begin{cases}
          φ₂ &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁, φ₂ ≡ φ₄,\\
          φ₂ ∨ φ₄ &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁,\\
          φ &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{theorem}[thm-resolve]
  \label{thm:thm-resolve}
  $Γ ⊢ φ₁$, $Γ ⊢ φ₂$, a literal $l$, and $ψ\in\Prop$ then $Γ ⊢ resolve(φ₁, φ₂, l, ψ)$, where,
  \begin{equation*}
    resolve(φ₁, φ₂, l, ψ) =
      rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨} (φ₂, ¬ l ∨ ψ)).
  \end{equation*}
\end{theorem}

\begin{proof}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ₁$}
    \RightLabel{thm-reorder$_{∨}$}
    \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ)$}

    \AxiomC{$Γ ⊢ φ₂$}
    \RightLabel{thm-reorder$_{∨}$}
    \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}

    \RightLabel{∧-intro}
    \BinaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
    \RightLabel{lem-rsol}
    \UnaryInfC{$Γ ⊢ rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ))$}
  \end{bprooftree}
  \end{equation*}
\end{proof}

% subsubsection resolve (end)
% --------------------------------------------------------------------

% --------------------------------------------------------------------

\subsubsection{Normal Forms.} % (fold)
\label{ssub:normal_forms}

A negative normal form for the formula φ is one in which negations
appear only in the literals and the expression is in terms only of
the boolean operators (¬, ∧, ∨).

To transform a formula φ to its equivalent \abbre{NNF} form, we
apply recursively the following equations. The function type for
\name{nnf} is $\Prop \to \Prop$.

\begin{lemma}[lem-nnf]
  \label{lem:lem-nnf}
  $Γ ⊢ φ$ then $Γ ⊢ nnf(φ)$, where,
  \begin{align*}
    \begin{split}
    \label{eq:nnf-definition}
    nnf(φ) =
      \begin{cases}
      nnf(φ₁) ∧ nnf(φ₂)
         &\text{if }φ ≡ φ₁ ∧ φ₂, \\
      nnf(φ₁) ∨ nnf(φ₂)
         &\text{if }φ ≡ φ₁ ∨ φ₂, \\
      nnf(¬ φ₁ ∨ φ₂)
         &\text{if }φ ≡ φ₁ ⇒ φ₂, \\
      nnf(φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁)
         &\text{if }φ ≡ φ₁ ⇔ φ₂, \\
      nnf(¬ φ₁ ∨ ¬ φ₂)
         &\text{if }φ ≡ ¬ (φ₁ ∧ φ₂), \\
      nnf(¬ φ₁ ∧ ¬ φ₂)
         &\text{if }φ ≡ ¬ (φ₁ ∨ φ₂), \\
      nnf(φ₁)
         &\text{if }φ ≡ ¬ ¬ φ₁, \\
      nnf(¬ (φ₂ ∨ ¬ φ₁))
         &\text{if }φ ≡ ¬ (φ₁ ⇒ φ₂), \\
      nnf(¬ (φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁))
         &\text{if }φ ≡ ¬ (φ₁ ⇔ φ₂), \\
      ⊥  &\text{if }φ ≡ ¬ ⊤, \\
      ⊤  &\text{if }φ ≡ ¬ ⊥, \\
      φ  & \text{otherwise}.
      \end{cases}
    \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-dist$_{∨}$]
  \label{lem:lem-dist-or}
  $Γ ⊢ φ$ then $Γ ⊢ dist_{∨}(φ)$, where,
  \begin{align*}
      \begin{split}
      dist_{∨} =
        \begin{cases}
          dist_{∨}(φ₁ ∨ φ₂) ∧ dist_{∨}(φ₂ ∨ φ₃)
            &\text{if }φ ≡ (φ₁ ∧ φ₂) ∨ φ₃,\\
          dist_{∨}(φ₁ ∨ φ₃) ∧ dist_{∨}(φ₁ ∨ φ₃)
            &\text{if }φ ≡ φ₁ ∨ (φ₂ ∧ φ₃),\\
          φ &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-dist]
  \label{lem:lem-dist}
  $Γ ⊢ φ$ then $Γ ⊢ dist(φ)$, where,
  \begin{align*}
      \begin{split}
      dist =
        \begin{cases}
          dist(φ₁) ∧ dist(φ₂)
            &\text{if }φ ≡ φ₁ ∧ φ₂,\\
          dist_{∨}(dist(φ₁) ∨ dist(φ₂))
            &\text{if }φ ≡ φ₁ ∨ φ₂,\\
          φ &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{proof} Sketch. Case when $φ ≡ φ₁ ∨ φ₂$.\\
For $i = 1, 2$,
\begin{equation}
  \begin{bprooftree}
    \RightLabel{assume}
    \AxiomC{$Γ, φ_{i} ⊢ φ_{i}$}
    \RightLabel{lem-dist}
    \UnaryInfC{$Γ ⊢ dist(φ_{i})$}
    \RightLabel{∨-intro$_{1,2}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist(φ₁) ∨ dist(φ₂)$}
    \RightLabel{lem-dist$_{∨}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$.}
    \end{bprooftree}
\end{equation}
Then, by using the ∨-elim rule, from the last derivation above we derive
$Γ, φ₁ ∨ φ₂ ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$, and the lemma follows.
\end{proof}

\begin{theorem}[thm-cnf]
\label{thm:thm-cnf}
  $Γ ⊢ φ$ then $Γ ⊢ cnf(φ)$, where,
  \begin{equation}
    cnf(φ) = (dist \circ nnf) (\varphi)
  \end{equation}
\end{theorem}

\begin{proof}
  Composition of the Lemma \ref{lem:lem-dist} and Lemma \ref{lem:lem-nnf}.
\end{proof}
% subsubsection normal_forms (end)
% --------------------------------------------------------------------

% --------------------------------------------------------------------

\subsubsection{Clausification.} % (fold)
\label{ssub:clausification}

\begin{theorem}[thm-clausify] % (fold)
\label{thm:thm_clausify}
  $Γ ⊢ φ$ and $ψ \in\Prop$ then $Γ ⊢ clausify(φ, ψ)$, where,
  \begin{align}
      clausify(φ,ψ) =
      \begin{cases}
        ψ, &\text{ if }φ≡ψ\\
        reorder_{∧∨}(cnf(φ), ψ), &\text{otherwise.}
      \end{cases}
  \end{align}
\end{theorem}

\begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~\ref{eq:substitution-theorem}. Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
\end{proof}

% subsubsection clausification (end)
% --------------------------------------------------------------------
